---
title: "data_transformations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{data_transformations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r load-packages}
library(adlgraphs)
library(tibble)
library(dplyr)
library(labelled)
library(haven)
library(waldo)
```


`adlgraphs` provides three main functions to reduce the amount of time it takes to perform a few very common data transformations:

- `num_rev()` - reverses the values of a vector

- `make_dicho()` - converts a vector to a dichotomous factor

- `make_binary()` - converts a vector to a binary (0 and 1)

## `num_rev()` 

`num_rev()` was designed with `forcats::fct_rev()` in mind. However, instead of operating on factors, `num_rev()` operates on numeric vectors. Often, when working with survey data, we want to reverse the values of a numeric vector. However, when this is done, any underlying metadata of the variable is lost. The purpose of `num_rev()` is to fix this by automate the process of reversing a numeric vector, while maintaining the variable and value labels. In addition, this function adds a new attribute called `transformation` that describes the data transformation used to create this variable. 

In this vignette, I hope to show you how this function makes life a lot easier if you want to reverse a variable while maintaining the metadata. The examples in this vignette get increasingly complex as I progress.

Let's look at an example with a fake data set comprised of four variables. In this data set, `w` has no value labels or variable label, `x` has only a variable label, `y` has only value labels, `z` has both variable and value labels. This allows us to see how the function provides different outputs depending on what kind of metadata the function has.


```{r}
library(adlgraphs)
library(tibble)
library(dplyr)
library(labelled)
library(haven)
library(waldo)

# create the new data
df <- tibble::tibble(
  w = sample(1:4, 20, replace = TRUE),
  x = sample(1:4, 20, replace = TRUE),
  y = sample(1:4, 20, replace = TRUE),
  z = sample(1:4, 20, replace = TRUE)
) %>% 
  # add value labels to y and z but not x to see what happens with no labels
  labelled::set_value_labels(
    z = c(`Strongly agree` = 1,
          `Somewhat agree` = 2,
          `Somewhat disagree` = 3,
          `Strongly disagree` = 4),
    y = c(`Strongly agree` = 1,
          `Somewhat agree` = 2,
          `Somewhat disagree` = 3,
          `Strongly disagree` = 4),
  ) %>%
  # set variable labels to x and z
  labelled::set_variable_labels(
    x = "This is the variable label for x",
    z = "This is the variable label for z"
  )
```

Now let's show how this function works in practice as well as how you would do the same data transformation without this function. We will start by showing it `w` because it has no metadata which makes it relatively easy to replicate without the function. 

```{r}
new_df <- df %>% 
  mutate(
    fun_w = num_rev(w),
    # first subtract w from 5 and then use structure to add the metadata
    base_w = 5 - w %>% 
      structure(transformation = "Reversing 'w'"),
    .keep = "used"
  )


# see the results
head(new_df)
```

Now let's check to see if `fun_w` and `base_w` are identical using `waldo::compare()`. We can see that there are no differences between the two vectors. In addition, check the underlying structure of the data, specifically of the two variables. We can see that both variables have an attribute called "transformation" that contains information explaining what happened sort of data transformation the variables underwent. 


```{r}
# use waldo::compare to see if there are differences
waldo::compare(new_df$fun_w, new_df$base_w)

# check the structure of the data now
str(new_df)
```

Now let's check this with `x` from the new data. Let's compare a new variable created using the `num_rev()` function called `num_rev_x` to a new variable created through a simple reversal by doing `5 - x` that we'll call . We can see that the values are the same, but hte 
```{r}
new_df <- df %>% 
  mutate(
    fun_x = num_rev(x),
    sub_x = 5 - x
  )
```


Since `x` has a variable label, it will be a bit more in-depth having to replicate the results we get from the `num_rev()` function. We can see that both variables are the same. Moreover, what was just one line of code ended up being ~5 lines of code. While this may not seem like much, it becomes a lot when you have a bunch of variables to do this over. Now to understand the output from `str(new_df)`, there are three things to note here:

 - The first is that the first line of each variable indicates the values in this variable and that there are 10 values.

 - The second line shows you the there is a variable label attribute

 - The third line shows you the transformation attribute

```{r}
new_df <- df %>% 
  mutate(
    fun_x = num_rev(x),
    # use haven labelled to add the variable label
    base_x = haven::labelled(
      # reverse the vector
      5 - x,
      # add the variable label
      label = "This is the variable label for x"
    ) %>% 
      # add the attribute for the transformation
      structure(transformation = "Reversing 'x'"),
    .keep = "used"
  )


# check to see if the data is the same
waldo::compare(new_df$fun_x, new_df$base_x)

# check the structure of the data now
str(new_df)
```

Now if we check `y` we will see something similar. First, since this variable has value labels, we need tomake sure that we add the value labels in when doing it in the base model and note that in the transformation. More importantly, we can see that while the values in `fun_y` and `base_y` changed, the value labels stayed the same as in `y`. This is important because the person's response didn't change, the only thing that did is the value we are associating with it. Moreover, we can that the "label" attribute is gone now and two "labels" attributes show up instead. This is because the original variable did not have a variabel label but did have value labels.

```{r}
new_df <- df %>% 
  mutate(
    fun_y = num_rev(y),
    # use haven labelled to add the variable label
    base_y = haven::labelled(
      # reverse the vector
      5 - y,
      # add the variable label
      labels = c("Strongly disagree" = 1,
                 "Somewhat disagree" = 2,
                 "Somewhat agree" = 3,
                 "Strongly agree" = 4)
    ) %>% 
      # add the attribute for the transformation
      structure(transformation = "Reversing 'y' while maintaining correct value labels"),
    .keep = "used"
  )

# preview data
head(new_df)

# check to see if the data is the same
waldo::compare(new_df$fun_y, new_df$base_y)

# check the structure of the data now
str(new_df)
```

`z` is the most complex variable as it contains both variable and value labels. What was originally one line of code quickly became 9 lines of code. To get the same result as `num_rev()`, we need to use `haven::labelled()` to add both the variable label and the value labels and we need to add `structure()` to add the transformation.

```{r}
new_df <- df %>% 
  mutate(
    fun_z = num_rev(z),
    # use haven labelled to add the variable label
    base_z = haven::labelled(
      # reverse the vector
      5 - z,
      # add the variable label
      labels = c("Strongly disagree" = 1,
                 "Somewhat disagree" = 2,
                 "Somewhat agree" = 3,
                 "Strongly agree" = 4),
      label = c("This is the variable label for z")
    ) %>% 
      # add the attribute for the transformation
      structure(transformation = "Reversing 'z' while maintaining correct value labels"),
    .keep = "used"
  )

# preview data
head(new_df)

# check to see if the data is the same
waldo::compare(new_df$fun_z, new_df$base_z)

# check the structure of the data now
str(new_df)
```












