---
title: "data_transformations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{data_transformations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r load-packages, message = FALSE}
# load in the libraries

# for num_rev and test_data
library(adlgraphs)
# this is for some other basic data transformations
library(dplyr)
# for working with labelled data
library(labelled)
# to make labelled data
library(haven)
# to check for differences
library(waldo)

```


`adlgraphs` provides three main functions to reduce the amount of time it takes to perform a few very common data transformations:

- `num_rev()` - reverses the values of a vector

- `make_dicho()` - converts a vector to a dichotomous factor

- `make_binary()` - converts a vector to a binary (0 and 1)

## `num_rev()` 

`num_rev()` was designed with `forcats::fct_rev()` in mind. However, instead of operating on factors, `num_rev()` operates on numeric vectors. Let's take a look at how this operates in practice. To do so, let's look at the variable `prod_isr` from the data set `test_data`. As we can see, survey respondents saw the a positively valanced statement, "I would feel comfortable buying products from Israel", with the values ranging from 1 to 4 where `1 = Strongly agree` and `4 = Strongly disagree`.

```{r}
str(test_data$prod_isr)
```

In survey research, we often have to reverse a variable. This is primarily done for two reasons. The first is because we want to flip the valence of the question. One reason this might be the case is because we wanted to create an index by summing this up with other variables. If the other statements were negatively valanced, it may be in our best interest to keep the value labels the same but reverse the valance of `prod_isr` so that it can now be interpreted as "I would not feel comfortable buying products from Israel." Since the value labels don't change, the people who were previously classified as "Strongly agree" are now going be classified as "Strongly disagree", etc.

The second reason one might reverse the values, is to change the direction of the scale. For example, as the values in `prod_isr` increase, so does one's level of disagreement. We might want to reverse the values so that now a 1 = "Strongly disagree" and a 4 = "Strongly agree", and thus a higher number means more agreement. Because this method does not flip the valance of the question, people who were originally classified as "Strongly agree" are still going to be classified as "Strongly agree". Put another way, the question isn't changing and therefore people's responses aren't changing. The only thing that changes is the value associated with their response.

`num_rev()` was designed with the second goal in mind. Normally when simply reversing the values by subtracting them, all underlying metadata and attributes are lost. As a result, we would have to reverse the values, update the value labels and set the variable label again. This is really time consuming. The purpose of `num_rev()` is to fix this by automating this process of reversing a numeric vector, while maintaining the variable and value labels. In addition, this function adds a new attribute called `transformation` that describes the data transformation used to create this variable. 

I'm now going to show this function in action. So we can see that while the values are flipped, the value labels for each respondent is the same.

```{r}
new_df <- test_data %>% 
  # let's make a new variable with the num_rev function
  mutate(prod_isr_rev = num_rev(prod_isr)) %>% 
  # keep only these two variables
  select(prod_isr_rev, prod_isr)

head(new_df)
```

As we can see below, the variable labels are also the same. This means that the only thing that really changed, is the values.

```{r}
test_data %>% 
  # let's make a new variable with the num_rev function
  mutate(prod_isr_rev = num_rev(prod_isr)) %>% 
  # keep only these two variables
  select(prod_isr_rev, prod_isr) %>% 
  # get the variable labels for both variables
  get_variable_labels()

```

Now that it's clear what this function does and what it doesn't do, let's try manually doing it. First, let's try just subtracting 5 from `prod_isr`. As we can see all attributes have been stripped.

```{r}
new_df <- test_data %>% 
  # reverse the variable
  mutate(prod_isr_rev = 5 - prod_isr)

# check attributes
attributes(new_df$prod_isr_rev)

```

This means we need to add both value and variable labels to the variable. To do this, we'll use the `haven::labelled()` function.

```{r}
new_df <- test_data %>% 
  mutate(
    prod_isr_rev = labelled(
      x = 5 - prod_isr,
      labels = c(
        `Strongly disagree` = 1,
        `Somewhat disagree` = 2,
        `Somewhat agree` = 3,
        `Strongly agree` = 4
      ),
      label = "I am comfortable buying products from Israel"
    )
  )

attributes(new_df$prod_isr_rev)

```

There is an easier of doing this by using `labelled::val_labels()` and `labelled::var_label()` to leverage the existing value and variable labels from `prod_isr`. However, when setting the value labels, we have to use `rev()` to reverse them.

```{r}
new_df <- test_data %>% 
  mutate(
    prod_isr_rev = labelled(
      x = 5 - prod_isr,
      labels = val_labels(test_data$prod_isr) %>% rev(),
      label = var_label(test_data$prod_isr)
    )
  )

attributes(new_df$prod_isr_rev)
```

Now that you know how to do it manually, I think it's clear that using `num_rev()` is much easier.





