---
title: "Data Transformations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Transformations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r load-packages, message = FALSE}
# load in the libraries

# for num_rev and test_data
library(adlgraphs)
# this is for some other basic data transformations
library(dplyr)
# for working with labelled data


```


`adlgraphs` provides three main functions to reduce the amount of time it takes to perform a few very common data transformations:

- `num_rev()` - reverses the values of a vector

- `make_dicho()` - converts a vector to a dichotomous factor

- `make_binary()` - converts a vector to a binary (0 and 1)

## `num_rev()` 

`num_rev()` was designed with `forcats::fct_rev()` in mind. However, instead of operating on factors, `num_rev()` operates on numeric vectors. Let's take a look at how this operates in practice. To do so, let's look at the variable `top` from the data set `test_data`. As we can see, survey respondents saw the a positively valanced statement, "I would feel comfortable buying products from Israel", with the values ranging from 1 to 4 where `1 = Strongly agree` and `4 = Strongly disagree`.

```{r}
str(test_data$top)
```

In survey research, we often have to reverse a variable. This is primarily done for two reasons. The first is because we want to flip the valence of the question. One reason this might be the case is because we wanted to create an index by summing this up with other variables. If the other statements were negatively valanced, it may be in our best interest to keep the value labels the same but reverse the valance of `top` so that it can now be interpreted as "I would not feel comfortable buying products from Israel." Since the value labels don't change, the people who were previously classified as "Strongly agree" are now going be classified as "Strongly disagree", etc.

The second reason one might reverse the values, is to change the direction of the scale. For example, as the values in `top` increase, so does one's level of disagreement. We might want to reverse the values so that now a 1 = "Strongly disagree" and a 4 = "Strongly agree", and thus a higher number means more agreement. Because this method does not flip the valance of the question, people who were originally classified as "Strongly agree" are still going to be classified as "Strongly agree". Put another way, the question isn't changing and therefore people's responses aren't changing. The only thing that changes is the value associated with their response.

`num_rev()` was designed with the second goal in mind. Normally when simply reversing the values by subtracting them, all underlying metadata and attributes are lost. As a result, we would have to reverse the values, update the value labels and set the variable label again. This is really time consuming. The purpose of `num_rev()` is to fix this by automating this process of reversing a numeric vector while maintaining the variable and value labels. In addition, this function adds a new attribute called `transformation` that describes the data transformation used to create this variable. 

I'm now going to show this function in action. We can see that when `top` = 1, `top_rev` = 4; when `top` = 2, `top_rev` = 3; `top` = 3, `top_rev` = 2; and when `top` = 4, `top_rev` = 1. 

```{r}
new_df <- test_data %>% 
  # let's make a new variable with the num_rev function
  mutate(top_rev = num_rev(top)) %>% 
  # keep only these two variables
  select(top_rev, top)

head(new_df)
```

This function does a lot more than just `5 - top`. Let's take a look. Using `str()` we can see that both variables have "label" and "labels" attributes and `top_rev` has a new attribute called "transformation". The new "transformation" attribute is automatically added and describes what sort of data transformation the variable underwent when it was created. This is valuable in case you forgot how it was created. In addition, both variables have the same variable label as seen in the "label" attribute. The key difference between these can be found in the "labels" attribute. In `top_rev`, the labels are reversed. In the original variable, `top`, 1 = "Strongly agree", 2 = "Somewhat agree", etc. However, in the new variable 4 = "Strongly agree", 3 = "Somewhat agree", etc.

```{r}

str(new_df)

```

Now, there are two reasons that the different labels is significant. First, it means that when we check the frequencies for both variables we will get the same results. The only difference is the order of the response options.

```{r}
get_freq_table(new_df, top_rev)

get_freq_table(new_df, top)

```

Second, the means of the variable will be different. For `top`, higher score means more disagreement. However, for `top_rev`, a higher number means more agreement. We can see the differences below.

```{r}

get_means(new_df, top_rev)

get_means(new_df, top)

```





